Thermodynamic response functions in a cell fluid model

### 2024-09-15 paper_text.tex
punctuation

Thermodynamic response functions play a crucial role in understanding and characterizing the behavior of physical systems. These functions describe how a system responds to changes in its state variables, providing valuable insights into its thermodynamic properties. The most widely studied thermodynamic response functions of fluids are isothermal and adiabatic compressibilities, isobaric thermal expansion, isochoric thermal pressure coefficient and heat capacities either at constant pressure or at constant volume. Thermodynamic response functions are essential tools for understanding, predicting, and controlling the behavior of physical systems. They are used in various scientific and engineering disciplines to model, design, and optimize processes and to gain insights into the fundamental principles governing thermodynamic systems. Therefore, a key area of research involves examining the thermodynamic properties of simple fluids and fluid models through theoretical approaches, computer simulations, and experimental studies, covering both subcritical and supercritical regions.~\cite{Johnston2014book,YigzaweSadus2013,Velasco2011,Bulavin2024}.

@@

In Section~\ref{sec:potential} we present the modified Morse potential used in our work, and briefly compare it with other possible modifications. Section~\ref{sec:eos} is dedicated to the explicit expressions for the equation of state. In Section~\ref{sec:respons_functions} the expressions for response functions are derived in terms of thermodynamic derivatives suitable for different forms of the equation of state.

@@

The function $E_\mu(M, T)$ from the equation \eqref{eq:eosMT} is provided by
\begin{eqnarray}\label{eq:E_mu}
	E_\mu (M, T) & = & - \frac{\ln (2\pi \beta W(0))}{2 N_v}  
	\nonumber\\
	&& +  g_0 - \frac{\beta W(0)}{2} 
	\left(\frac{\tilde\mu}{W(0)} \right)^{2} 
	\nonumber\\
	&& - \frac{g_3}{g_4} {M} \! - \frac{g_3^2}{2 g_4^2}  d - \frac{1}{24} \frac{g_3^4}{g_4^3}. 
\end{eqnarray}
Here the quantity $N_v$ defines the number of cubic cells in volume $V$ for the initial model.
In the thermodynamic limit, $N_v \to \infty$, and thus the first term can be neglected. The term $\tilde{\mu}/W(0)$ can be expressed in terms of $M$ using~\eqref{chem_pot}. The temperature and the inverse temperature can always be expressed in terms of the reduced temperature and a corresponding critical value:
\begin{equation*}
	T = T_c(1+\tau), \quad \beta = \beta_c (1 + \tau)^{-1}
\end{equation*}

@@

\subsection{Isothermal compressibility}
The isothermal compressibility is defined by
\begin{equation}
	\label{def:isotherm_compres}
	\kappa_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T, N}.
\end{equation}
Let us perform some transformations to rewrite $\kappa_T$ into a form that is more suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}:
\begin{eqnarray*}
	\kappa_T	& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,N}
	\\
	& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T}}
	{\left(\partial P / \partial \mu\right)_{T}}.
\end{eqnarray*}
We have omitted the condition of constant $N$ in the last line of the above equation since we have explicit dependencies on temperature and chemical potential for both pressure $P = P(T, \mu)$, Eq.~\eqref{eq:eosMT}, and density $\rho = \rho(T, \mu)$, Eq.~\eqref{eq:density}.
Applying the Gibbs--Duhem equation
\begin{equation}
	N{\rm d}\mu = -S{\rm d}T + V{\rm d}P,
\end{equation}
at $T = const$ one has 
\begin{equation*}
	{\rm d} P = \rho {\rm d} \mu,
\end{equation*}
or
\begin{equation}
	\label{eq:rho_dpdm}
	\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
\end{equation}
Substituting this into the last expression for $\kappa_T$, one gets
\begin{equation}
	\kappa_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T}
\end{equation}
Finally, from~\eqref{eq:rho_dpdm} it follows that
\begin{equation}
	\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T
\end{equation}
and ultimately we arrive at the very useful expression for the isothermal compressibility
\begin{equation}
	\kappa_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T}
\end{equation}

@@

\subsection{Thermal pressure coefficient}
The thermal pressure coefficient is defined by
\begin{equation}
	\label{def:therm_pres_coef}
	\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,N}.
\end{equation}
We rewrite $\beta_V$ into a form that is suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}
\begin{eqnarray}
	\beta_V	& = & \left(\frac{\partial P}{\partial T}\right)_{\mu} + \left(\frac{\partial P}{\partial \mu}\right)_T \left(\frac{\partial \mu}{\partial T}\right)_{V, N}.
\end{eqnarray}
Applying the cyclic relation
\begin{equation*}
	\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
	\left(\frac{\partial T}{\partial V}\right)_{\mu, N}
	\left(\frac{\partial V}{\partial \mu}\right)_{T, N}
	= -1
\end{equation*}
we obtain
\begin{eqnarray*}
	\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
	& = & 
	- \left(\frac{\partial V}{\partial T}\right)_{\mu, N} 
	\left(\frac{\partial V}{\partial \mu}\right)^{-1}_{T, N} \\
	& = & - \left(\frac{\partial \rho^*}{\partial T}\right)_{\mu}
	\left(\frac{\partial \rho^*}{\partial \mu}\right)^{-1}_{T}.		
\end{eqnarray*}
Substituting this result into the formula for $\beta_V$, we arrive at the final expression for the thermal pressure coefficient
\begin{equation}
	\beta_V = \left(\frac{\partial P}{\partial T}\right)_{\mu} 
	- \left(\frac{\partial P}{\partial \mu}\right)_T 
	\left(\frac{\partial \rho^*}{\partial T}\right)_{\mu}
	\left(\frac{\partial \rho^*}{\partial \mu}\right)^{-1}_{T}
\end{equation}
which\footnote{Compare this equation for $\beta_V$ with Eq.(17) from~\cite{StrokerMeier2021}} is easily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}. It is also worth noting that the first contribution to $\beta_V$ is essentially the entropy per volume, $S/V = (\partial P / \partial T)_\mu$.

@@

If it is preferable to use the equation of state in the form $P=P(T, \rho^*)$, see~\eqref{eq:eosNT} and~\eqref{eq:eosPTn_reduced}, then we need to rewrite the definition for $\alpha_P$ accordingly. We start with~\eqref{eq:alpha_vs_rho} and apply the cyclic relation
\begin{equation}
	\left(\frac{\partial \rho}{\partial T}\right)_P = - \left(\frac{\partial P}{\partial T}\right)_{\rho} \left(\frac{\partial P}{\partial \rho}\right)^{-1}_T
\end{equation}
to get for $\alpha_P$
\begin{equation}
	\alpha_P = \frac{1}{\rho} \left(\frac{\partial P}{\partial T}\right)_{\rho} \left(\frac{\partial P}{\partial \rho}\right)^{-1}_T.
\end{equation}
or for $\alpha^*_P$
\begin{equation}
	\alpha^*_P = \frac{1}{\rho^* T^*_c} \left(\frac{\partial P^*}{\partial \tau}\right)_{\rho} \left(\frac{\partial P^*}{\partial \rho^*}\right)^{-1}_T.
\end{equation}

>>>>

Thermodynamic response functions play a crucial role in understanding and characterizing the behavior of physical systems. These functions describe how a system responds to changes in its state variables, providing valuable insights into its thermodynamic properties. The most widely studied thermodynamic response functions of fluids are isothermal and adiabatic compressibilities, isobaric thermal expansion, isochoric thermal pressure coefficient and heat capacities either at constant pressure or at constant volume. Thermodynamic response functions are essential tools for understanding, predicting, and controlling the behavior of physical systems. They are used in various scientific and engineering disciplines to model, design, and optimize processes and to gain insights into the fundamental principles governing thermodynamic systems. Therefore, a key area of research involves examining the thermodynamic properties of simple fluids and fluid models through theoretical approaches, computer simulations, and experimental studies, covering both subcritical and supercritical regions~\cite{Johnston2014book,YigzaweSadus2013,Velasco2011,Bulavin2024}.

@@

In Section~\ref{sec:potential}, we present the modified Morse potential used in our work, and briefly compare it with other possible modifications. Section~\ref{sec:eos} is dedicated to the explicit expressions for the equation of state. In Section~\ref{sec:respons_functions}, the expressions for response functions are derived in terms of thermodynamic derivatives suitable for different forms of the equation of state.

@@

The function $E_\mu(M, T)$ from the equation \eqref{eq:eosMT} is provided by
\begin{eqnarray}\label{eq:E_mu}
	E_\mu (M, T) & = & - \frac{\ln (2\pi \beta W(0))}{2 N_v}  
	\nonumber\\
	&& +  g_0 - \frac{\beta W(0)}{2} 
	\left(\frac{\tilde\mu}{W(0)} \right)^{2} 
	\nonumber\\
	&& - \frac{g_3}{g_4} {M} \! - \frac{g_3^2}{2 g_4^2}  d - \frac{1}{24} \frac{g_3^4}{g_4^3}. 
\end{eqnarray}
Here the quantity $N_v$ defines the number of cubic cells in volume $V$ for the initial model.
In the thermodynamic limit, $N_v \to \infty$, and thus, the first term can be neglected. The term $\tilde{\mu}/W(0)$ can be expressed in terms of $M$ using~\eqref{chem_pot}. The temperature and the inverse temperature can always be expressed in terms of the reduced temperature and a corresponding critical value:
\begin{equation*}
	T = T_c(1+\tau), \quad \beta = \beta_c (1 + \tau)^{-1}.
\end{equation*}

@@

\subsection{Isothermal compressibility}
The isothermal compressibility is defined by
\begin{equation}
	\label{def:isotherm_compres}
	\kappa_T = -\frac{1}{V}\left(\frac{\partial V}{\partial P}\right)_{T, N}.
\end{equation}
Let us perform some transformations to rewrite $\kappa_T$ into a form that is more suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced}:
\begin{eqnarray*}
	\kappa_T	& = & \frac{1}{\rho} \left(\frac{\partial \rho}{\partial P} \right)_{T,N}
	\\
	& = & \frac{1}{\rho} \frac{\left(\partial \rho / \partial \mu\right)_{T}}
	{\left(\partial P / \partial \mu\right)_{T}}.
\end{eqnarray*}
We have omitted the condition of constant $N$ in the last line of the above equation since we have explicit dependencies on temperature and chemical potential for both pressure $P = P(T, \mu)$, Eq.~\eqref{eq:eosMT}, and density $\rho = \rho(T, \mu)$, Eq.~\eqref{eq:density}.
Applying the Gibbs--Duhem equation
\begin{equation}
	N{\rm d}\mu = -S{\rm d}T + V{\rm d}P,
\end{equation}
at $T = const$ one has 
\begin{equation*}
	{\rm d} P = \rho {\rm d} \mu,
\end{equation*}
or
\begin{equation}
	\label{eq:rho_dpdm}
	\rho = \left(\frac{\partial P}{\partial \mu}\right)_T.
\end{equation}
Substituting this into the last expression for $\kappa_T$, one gets
\begin{equation}
	\kappa_T = \frac{1}{\rho^2} \left(\frac{\partial \rho}{\partial \mu}\right)_{T} .
\end{equation}
Finally, from~\eqref{eq:rho_dpdm} it follows that
\begin{equation}
	\left(\frac{\partial \rho}{\partial \mu}\right)_T = \left(\frac{\partial^2 P}{\partial \mu^2}\right)_T ,
\end{equation}
and ultimately we arrive at the very useful expression for the isothermal compressibility
\begin{equation}
	\kappa_T = \frac{1}{\rho^2} \left(\frac{\partial^2 P}{\partial \mu^2}\right)_{T} .
\end{equation}

@@

\subsection{Thermal pressure coefficient}
The thermal pressure coefficient is defined by
\begin{equation}
	\label{def:therm_pres_coef}
	\beta_V = \left( \frac{\partial P}{\partial T} \right)_{V,N}.
\end{equation}
We rewrite $\beta_V$ into a form that is suitable for the equation of state $P=P(T,\mu)$, see~\eqref{eq:eosMT} and~\eqref{eq:eosPTM_reduced},
\begin{eqnarray}
	\beta_V	& = & \left(\frac{\partial P}{\partial T}\right)_{\mu} + \left(\frac{\partial P}{\partial \mu}\right)_T \left(\frac{\partial \mu}{\partial T}\right)_{V, N}.
\end{eqnarray}
Applying the cyclic relation
\begin{equation*}
	\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
	\left(\frac{\partial T}{\partial V}\right)_{\mu, N}
	\left(\frac{\partial V}{\partial \mu}\right)_{T, N}
	= -1
\end{equation*}
we obtain
\begin{eqnarray*}
	\left(\frac{\partial \mu}{\partial T}\right)_{V, N} 
	& = & 
	- \left(\frac{\partial V}{\partial T}\right)_{\mu, N} 
	\left(\frac{\partial V}{\partial \mu}\right)^{-1}_{T, N} \\
	& = & - \left(\frac{\partial \rho^*}{\partial T}\right)_{\mu}
	\left(\frac{\partial \rho^*}{\partial \mu}\right)^{-1}_{T}.		
\end{eqnarray*}
Substituting this result into the formula for $\beta_V$, we arrive at the final expression for the thermal pressure coefficient
\begin{equation}
	\beta_V = \left(\frac{\partial P}{\partial T}\right)_{\mu} 
	- \left(\frac{\partial P}{\partial \mu}\right)_T 
	\left(\frac{\partial \rho^*}{\partial T}\right)_{\mu}
	\left(\frac{\partial \rho^*}{\partial \mu}\right)^{-1}_{T}
\end{equation}
which\footnote{Compare this equation for $\beta_V$ with Eq.(17) from~\cite{StrokerMeier2021}} is easily calculated based on Eqs.~\eqref{eq:eosMT} and~\eqref{eq:density}. It is also worth noting that the first contribution to $\beta_V$ is essentially the entropy per volume, $S/V = (\partial P / \partial T)_\mu$.

@@

If it is preferable to use the equation of state in the form $P=P(T, \rho^*)$, see~\eqref{eq:eosNT} and~\eqref{eq:eosPTn_reduced}, then we need to rewrite the definition for $\alpha_P$ accordingly. We start with~\eqref{eq:alpha_vs_rho} and apply the cyclic relation
\begin{equation}
	\left(\frac{\partial \rho}{\partial T}\right)_P = - \left(\frac{\partial P}{\partial T}\right)_{\rho} \left(\frac{\partial P}{\partial \rho}\right)^{-1}_T
\end{equation}
to get for $\alpha_P$
\begin{equation}
	\alpha_P = \frac{1}{\rho} \left(\frac{\partial P}{\partial T}\right)_{\rho} \left(\frac{\partial P}{\partial \rho}\right)^{-1}_T ,
\end{equation}
or for $\alpha^*_P$
\begin{equation}
	\alpha^*_P = \frac{1}{\rho^* T^*_c} \left(\frac{\partial P^*}{\partial \tau}\right)_{\rho} \left(\frac{\partial P^*}{\partial \rho^*}\right)^{-1}_T.
\end{equation}
